miniKanren with fair search strategies

### 2019-05-05 Paper/paper-slatex.tex
revise intro; revise fairness - fair disj

@@

\lstset{
  language=Scheme,
  basicstyle=\ttfamily,
  morekeywords={run,conde,run*,defrel,==,fresh},
  alsodigit={!\$\%&*+-./:<=>?@^_~},
  morecomment=[l]{;,\#{}lang},
  mathescape=true
}
%% scheme-list .

@@

\newcommand{\conde}{\texttt{cond$^e$} }
\newcommand{\conj}{\texttt{conj}}
\newcommand{\disj}{\texttt{disj}}
\newcommand{\clisting}[1]{
\begin{center}
  \begin{tabular}{c}
	\lstinputlisting{#1}
    \end{tabular}
\end{center}
}
\newcommand{\DFSi}[0]{
  DFS$_\textrm{i}$
}

@@

@@

The syntax of a programming language should reflect its semantics. When writing 
a \conde expression in miniKanren, a programmer would expect all clauses share 
the same chance of being explored, as these clauses are written in 
parallel. The existing search strategy, interleaving DFS (DFS$_\textrm{i}$), 
however, prioritize its clauses by the order how they are written down. 
Similarly, when a \conde is followed by another goal conjunctively, a 
programmer would expect answers in parallel share the same chance of being 
explored. Again, the answers by DFS$_\textrm{i}$ is different from the 
expectation. We have devised three new search strategies that have different 
level of fairness in \disj and \conj.

%The syntax of a programming language should reflect its semantics. When using 
%a disjunctive operator in relational programming, a programmer would expect 
%all 
%sub-goals of this disjunct to share the same chance of being explored, as 
%these 
%clauses are written in parallel. The existing multi-arity disjunctive operator 
%in miniKanren, however, prioritize its clauses by the order of which these 
%clauses are written down. We have devised two new search strategies that 
%allocate computational effort more fairly in all clauses.

@@

miniKanren is a family of relational programming languages.
\citet{friedman_reasoned_2018} introduce miniKanren and its implementation
in the The Reasoned Schemer, 2nd Ed(TR2). miniKanren programs, especially
relational interpreters, have been proven to be 
useful in solving many problems by \citet{byrd2017unified}.

\conde is the flagship disjunction operator in miniKanren. A subtlety arises
when a \conde contains many disjunction clauses: not every clause has the equal
chance to contribute to the result. As an example, consider the following
relation \texttt{repeato} and its invocation.
\clisting{Figures/repeato.rkt}
Next, consider the following disjuction of invoking \texttt{repeato} with
four different letters.
\clisting{Figures/example.rkt}
\conde intuitively corresponds to \texttt{or} in logic. And thus an unsuspicious
beginner would expect each letter to contribute equally to the result, as
following.
\clisting{Figures/run-repeato-fair.rkt}
The \conde in TR2, however, generates a less expected result.
\clisting{Figures/run-repeato-idfs.rkt}

\conde implements interleaving DFS(\DFSi), the cause of this unexpected result.
In a \conde, each clause takes a half of its received computational resources
and pass the other half to its following clauses, except for the last clause
that takes all resources it receives. In the example above, the letters
\texttt{c} and \texttt{d} are allocated with less resources than \texttt{a},
and thus \texttt{c} and \texttt{d} barely contribute to the result.

\DFSi is powerful for an expert. By carefully organizing the order of \conde
clauses, a miniKanren program can explore more ``interesting'' clauses than
those uninteresting ones, and thus use computational resources efficiently.
A little miniKanrener, however, may beg to differ--understanding
implementation details and fiddling with clauses order is not the first
priority of a beginner.

There is another reason that miniKanren could use more search strategies than
just \DFSi. In many applications, there does not exist one order that serves
for all purposes. For example, a relational dependent type checker contains
clauses for constructors that build data and clauses for eliminators that use
data. When this type checker is used to generate simple and shallow programs,
the clauses of constructors should be put in the front of \conde.
When performing proof search for complicated programs, the clauses of eliminator
should take the focus. With \DFSi, these two uses can not be efficient at the
same time. In fact, to make one use efficient, the other one must be drastically
slow.

\NOTE[LKC]{Should I explain ``state'' and ``search space'' here?}

Under the hood, \conde clauses are combined with \disj{}. The \disj{} 
implementation in \citep{friedman_reasoned_2018} is not \emph{fair}.

\begin{defn}[fair \disj]
	A \disj{} is fair iff it allocates computational resource evenly among 
	search spaces derived from disjunctive goals. \citet{seres1999algebra}
\end{defn}

A closely related concept is \emph{almost-fair} \disj.

\begin{defn}[almost-fair \disj]
	A \disj{} is almost-fair iff it allocates computational resource so evenly 
	among search spaces derived from disjunctive goals that the maximal ratio 
	of resources is bounded by a constant.
\end{defn}

The \disj{} in \citet{friedman_reasoned_2018}, or more precisely, in its search 
strategy, interleaving DFS(DFS$_\textrm{i}$), is neither fair nor almost-fair. 
Fair depth-first search (DFS$_\textrm{f}$), a new search strategy in this 
paper, has fair \disj. And balanced interleaving depth-first search 
(DFS$_\textrm{bi}$), another new strategy, has almost-fair \disj. Breath-first 
search (BFS), a known strategy \citet{seres1999algebra}, has fair \disj{} and 
\emph{fair \conj{}}.

\begin{defn}[fair \conj]
	A \conj{} is fair iff it allocates computational resource evenly among 
	search spaces derived from states in the same bag, where bags are finite 
	list of states.
\end{defn}

A comparison of fairness of search strategies is in Fig.~\ref{fairness}.

To summarize our contribution, we
\begin{itemize}
	\item propose a new concept, almost-fair \disj{}.
	\item propose a new definition of fair \conj{}.
	\item propose and implement balanced interleaving depth-first search 
	(DFS$_\textrm{bi}$), a new search strategy with almost-fair \disj{}.
	\item propose and implement fair depth-first search (DFS$_\textrm{f}$), a 
	new search strategy with fair \disj{}.
	\item implement in a new way breath-first search (BFS), a search strategy 
	with fair \disj{} and fair \conj{}
	(our code runs faster in all benchmarks and is simpler)
	\item prove our BFS implementation is equivalent with the one by 
	\citet{seres1999algebra}.
\end{itemize}

\begin{figure}[tbp]
	\begin{tabular}{|c|c|c|c|c|}
		\hline 
		fairness & DFS$_\textrm{i}$ & DFS$_\textrm{bi}$ & DFS$_\textrm{f}$ & BFS \\ 
		\hline 
		disj & unfair & almost-fair & fair & fair \\ 
		\hline 
		conj & unfair & unfair & unfair & fair \\ 
		\hline 
	\end{tabular} 
	\caption{fairness of search strategies}
	\label{fairness}
\end{figure}

% Interleaving DFS, the search strategy of \conde in
% miniKanren~\citet{friedman_reasoned_2018}, allocates computational resource
% by the order in which the clauses are written. Each clause of a
% \conde takes half of the current resource and passes the other half
% to its following clauses, except for the last clause that takes all of the 
%current resource. 
% The biased treatment provides both opportunity and burden: miniKanren users 
%can place more frequently used 
% goal at the beginning to optimize their programs; however, it might be a 
% catastrophe if a goal that generates many useless states is placed before 
%more 
% important goals. Seasoned miniKanreners usually know how to utilize 
% the unfairness to optimize their programs. However, we believe search 
%strategies that is less sensitive to goal order can also be useful to little 
%miniKanreners as well as seasoned ones. We propose two such search strategies, 
%balanced interleaving DFS (DFS$_\textrm{bi}$) and breadth-first search (BFS), and observe 
%how they affect the efficiency and the answer order of known miniKanren 
%programs. The experiment is conducted with the miniKanren from \textit{The 
%Reasoned Schemer, 2nd Edition}.

@@

In this section, we elaborate fairness by running queries about 
\texttt{repeato2}, a relational definition that relates a term \texttt{x} with a 
non-empty list whose elements are \texttt{x} (Fig.~\ref{repeato}).

\begin{figure}
	\lstinputlisting{Figures/repeato.rkt}
	\caption{\texttt{repeato} and an example run}
	\label{repeato2}
\end{figure}

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the answer. DFS$_\textrm{i}$, the current search
strategy, however, results in many more lists of \texttt{a}s than lists
of other letters. And some letters, e.g. \texttt{c} and \texttt{d}, are
rarely seen. The situation would be exacerbated if \conde contains more clauses.

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-disj-iDFS.rkt}
	\end{tabular}
\end{center}

We borrow the definition of \emph{fair} \disj{} from \citet{seres1999algebra}: 
search strategies with \emph{fair} \disj{} should allocate resources evenly 
among disjunctive goals. Running the same program with DFS$_\textrm{f}$ and BFS 
give the following result.

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-disj-fair.rkt}
	\end{tabular}
\end{center}

Now we are in a middle place between fair and unfair -- search strategies
with  \emph{almost-fair} \disj{} should allocate resources so evenly among
disjunctive goals that the maximal ratio of resources is bounded by a constant. 
Our new search strategy,
DFS$_\textrm{bi}$, has almost-fair \disj{}. It is
fair when the number of goals is a power of 2, otherwise, some goals are 
allocated twice as many resources than the others. In the previous example,
DFS$_\textrm{bi}$
gives the same result. And in the following example, where the \conde has 5 
clause, the clauses of \texttt{b}, \texttt{c}, and \texttt{d} are allocated 
more resources.

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-disj-biDFS.rkt}
	\end{tabular}
\end{center}

@@

In the following program, the three \conde clauses differ in a trivial way. So 
we expect lists of each letter constitute $1/4$ of the answer list. Search 
strategies with unfair \conj{} (e.g. DFS$_\textrm{i}$, DFS$_\textrm{f}$), 
however, give us many more lists of \texttt{a}s than lists of other letters. 
And some letters (e.g. lists 
of \texttt{c}) are rarely found. Although DFS$_\textrm{i}$'s \disj{} is unfair in general, 
it is fair when there is no call to relational definition in sub-goals, 
including this case. The situation would be worse if we add more \conde 
clauses. The result with DFS$_\textrm{bi}$, whose \conj{} is also unfair, is similar, but 
due to its different \disj{}, the position of \texttt{b} and \texttt{c} are 
swapped.

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-conj-unfair.rkt}
	\end{tabular}
\end{center}

Intuitively, search strategies with fair \conj{} should produce each letter of 
lists equally frequently. Indeed, BFS does so.

\begin{center}
	\begin{tabular}{c|c}
		\lstinputlisting{Figures/repeato-conj-fair.rkt}
	\end{tabular}
\end{center}

A more interesting situation is when the first conjunctive goal produces 
infinite many answers. Consider the following example, a naive specification of 
fair \conj{} might require search strategies to produce all sorts of singleton 
lists, but no longer ones, which makes the strategies \emph{incomplete}. 
\NOTE[MVC]{incomplete w.r.t. what? where's the definition of incomplete?}
\NOTE[LKC]{I am a bit confused. I assume completeness is a well-known concept 
in the context of logic programming. For example, this paper doesn't cite any 
source when it talks about completeness.

@@

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-conj-infinite-naive.rkt}
	\end{tabular}
\end{center}

Our solution requires a search strategy with \emph{fair} \conj{} to 
package answers in bags, where each bag contains finite answers, and to allocate 
resources evenly among search spaces derived from answers in the same bag. The 
way to package depends on search strategy. And how to allocate resources among 
search space related to different bags is unspecified. Our definition of fair 
\conj{} is orthogonal with completeness. For example, a naively fair strategy 
is fair but not complete, while BFS is fair and complete. 
\NOTE[MVC]{Also here, complete w.r.t what?}

@@

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-conj-infinite-sof.rkt}
	\end{tabular}
\end{center}

@@

\begin{figure}
	\lstinputlisting{Figures/balanced-disj.rkt}
	\caption{DFS$_\textrm{bi}$ implementation}
	\label{balanced-disj}
\end{figure}

Balanced interleaving DFS (DFS$_\textrm{bi}$) has almost-fair \disj{} and unfair \conj{}. 
The implementation of DFS$_\textrm{bi}$ differs from DFS$_\textrm{i}$ in the \disj{} macro. We list the 
new \disj{} with its helpers in Fig.~\ref{balanced-disj}. The first helper 
function, \texttt{split}, takes a list of goals \texttt{ls} and a procedure 
\texttt{k}, partitions \texttt{ls} into two sub-lists of roughly equal length, 
and returns the application of \texttt{k} to the two sub-lists. \texttt{disj*} 
takes a non-empty list of goals \texttt{gs} and returns a goal. With the help 
of \texttt{split}, it essentially constructs a \emph{balanced} binary tree 
where leaves are elements of \texttt{gs} and nodes are \texttt{disj2}s, hence 
the name of this search strategy. In contrast, the \disj{} in DFS$_\textrm{i}$
constructs the binary tree with the same nodes but in the unbalanced form.

@@

\begin{figure}
	\lstinputlisting{Figures/fDFS.rkt}
	\caption{DFS$_\textrm{f}$ implementation}
	\label{fDFS}
\end{figure}

Fair DFS (DFS$_\textrm{f}$) has fair \disj{} and unfair \conj{}. The implementation of DFS$_\textrm{f}$ 
differs from DFS$_\textrm{i}$'s in \texttt{disj2} (Fig.~\ref{fDFS}). \texttt{disj2} is 
changed to call a new and fair version of \texttt{append-inf}. 
\texttt{append-inf/fair} immediately calls its helper, 
\texttt{append-inf/fair\^{}}, with the first argument, \texttt{s?}, set to 
\texttt{\#{}t}, which indicates that \texttt{s-inf} and \texttt{t-inf} haven't 
been swapped. The swapping happens at the third \texttt{cond} clause in the 
helper, where \texttt{s?} is updated accordingly. The first two \texttt{cond} 
clauses essentially copy the \texttt{car}s and stop recursion when one of the 
input spaces is obviously finite. The third clause, as we mentioned above, is 
just for swapping. When the fourth and last clause runs, we know that both 
\texttt{s-inf} and \texttt{t-inf} are ended with a thunk. In this case, a new 
thunk is constructed. The new thunk calls the driver recursively. Here changing 
the order of \texttt{t-inf} and \texttt{s-inf} won't hurt the fairness (though 
it will change the order of answers). We swapped them back so that answers are 
produced in a more natural order.

@@

BFS is fair in both \disj{} and \conj{}. Our implementation is based on 
DFS$_\textrm{f}$ (not DFS$_\textrm{i}$). All we have to do is apply two trivial changes to 
\texttt{append-map-inf}. First, rename it to \texttt{append-map-inf/fair}. 
Second, replace its use of \texttt{append-inf} to \texttt{append-inf/fair}.

The implementation can be improved in two ways. First, as mentioned in 
section 2.2, BFS puts answers in bags and answers of the same cost are in the 
same bag. In this implementation, however, it is unclear where this information 
is recorded. Second, 
\texttt{append-inf/fair} is extravagant in memory usage. It makes 
$O(n+m)$ new \texttt{cons} cells every time, where $n$ and $m$ are the 
``length''s of input search spaces. We address these issues in the first 
subsection.

@@

\begin{figure}
	\lstinputlisting{Figures/BFS-opt.rkt}	
	\caption{new and changed functions in optimized BFS that implements pure 
	features}
	\label{BFS-opt}
\end{figure}

@@

As mentioned in section 2.2, BFS puts answers in bags and answers of the 
same cost are in the same bag. The cost
information is recorded subtly -- the \texttt{car}s of a search space have cost 
0 (i.e. they are in the same bag), and the costs of answers in thunk are 
computed recursively then increased by one. It is even more subtle that
\texttt{append-inf/fair} and the \texttt{append-map-inf/fair} respects the cost 
information. We make these facts more obvious by changing the type of search 
space, modifying related function definitions, and introducing a few more 
functions.

The new type is a pair whose \texttt{car} is a list of answers (the bag), and 
whose \texttt{cdr} is either a \texttt{\#{}f} or a thunk returning a search 
space. A falsy \texttt{cdr} means the search space is obviously finite.

Functions related to the pure subset are listed in Fig.~\ref{BFS-opt} (the 
others in Fig.~\ref{BFS-opt-cont}). They are compared with 
\citeauthor{seres1999algebra}'s implementation later. The first three functions 
in Fig.~\ref{BFS-opt} are search space constructors. \texttt{none} makes an 
empty search space; \texttt{unit} makes a space from one answer; and 
\texttt{step} makes a space from a thunk. The remaining functions do the same 
thing as before.

Luckily, the change in \texttt{append-inf/fair} also fixes the miserable space 
extravagance -- the use of \texttt{append} helps us to reuse the first bag of 
\texttt{t-inf}.

\citet{kiselyov2005backtracking} has shown that a \emph{MonadPlus} hides in 
implementations of logic programming system. Our BFS implementation is not an 
exception: \texttt{none}, \texttt{unit}, \texttt{append-map-inf}, and 
\texttt{append-inf} correspond to \texttt{mzero}, \texttt{unit}, \texttt{bind}, 
and \texttt{mplus} respectively.

\begin{figure}
	\lstinputlisting{Figures/BFS-opt-cont.rkt}	
	\caption{new and changed functions in optimized BFS that implements impure 
		features}
	\label{BFS-opt-cont}
\end{figure}

Functions implementing impure features are in Fig.~\ref{BFS-opt-cont}. The 
first function, \texttt{elim}, takes a space \texttt{s-inf} and two 
continuations \texttt{ks} and \texttt{kf}. When \texttt{s-inf} contains some 
answers, \texttt{ks} is called with the first answer and the rest space. 
Otherwise, \texttt{kf} is called with no argument. Here `s' and `f' means 
`succeed' and `fail' respectively. This function is an eliminator of search 
space, hence the name. The remaining functions do the same thing as before.

@@

We prove in Coq that the two BFSs are equivalent, i.e. \texttt{(run n g)} 
produces the same result (See supplements for the formal proof).

>>>>

%% slatex
\usepackage{slatex}
\setkeyword{defrel conde fresh run}
\setspecialsymbol{==}{$\equiv$}
\setspecialsymbol{conde}{\textbf{cond$^e$}}
\setspecialsymbol{append-inf}{append$^\infty$}
\setspecialsymbol{append-inf/fair}{append$^\infty_f$}
\setspecialsymbol{append-inf/fair^}{append$^\infty_f$\textasciicircum}
\setspecialsymbol{append-map-inf}{append-map$^\infty$}
\setspecialsymbol{append-map-inf/fair}{append-map$^\infty_f$}
\setspecialsymbol{repeato}{repeat$^o$}
\setspecialsymbol{appendo}{append$^o$}
\setspecialsymbol{very-recursiveo}{very-recursive$^o$}
\setspecialsymbol{g1}{$g_1$}
\setspecialsymbol{g2}{$g_2$}
\setspecialsymbol{g3}{$g_3$}
\setspecialsymbol{gs1}{$gs_1$}
\setspecialsymbol{gs2}{$gs_2$}
\setspecialsymbol{ks}{$k_s$}
\setspecialsymbol{kf}{$k_f$}
\setspecialsymbol{s-inf}{s$^\infty$}
\setspecialsymbol{t-inf}{t$^\infty$}
\setspecialsymbol{prod}{$\times$}
\setspecialsymbol{->}{$\rightarrow$}
\setspecialsymbol{#|}{\texttt{\#{}|}}
\setspecialsymbol{|#}{\texttt{|\#{}}}
\def\variablefont#1{{\tt#1\/}} 
\def\datafont#1{{\tt#1\/}} 
\def\constantfont#1{{\tt#1\/}}

%% scheme-list configuration
\lstset{
	language=Scheme,
	basicstyle=\ttfamily,
	morekeywords={run,conde,run*,defrel,==,fresh},
	alsodigit={!\$\%&*+-./:<=>?@^_~},
	morecomment=[l]{;,\#{}lang},
	mathescape=true
}

@@

@@

@@

% abbreviations
\newcommand{\dfsi}[0]{DFS$_\textrm{i}$}
\newcommand{\dfsbi}[0]{DFS$_\textrm{bi}$}
\newcommand{\dfsf}[0]{DFS$_\textrm{f}$}

@@

The syntax of a programming language should reflect its semantics. When writing 
a \scheme|conde| expression in miniKanren, a programmer would expect all clauses 
share the same chance of being explored, as these clauses are written in 
parallel. The existing search strategy, interleaving DFS (DFS$_\textrm{i}$), 
however, prioritize its clauses by the order how they are written down. 
Similarly, when a \scheme|conde| is followed by another goal conjunctively, a 
programmer would expect answers in parallel share the same chance of being 
explored. Again, the answers by DFS$_\textrm{i}$ is different from the 
expectation. We have devised three new search strategies that have different 
level of fairness in \scheme|disj| and \scheme|conj|.

@@

miniKanren is a family of relational programming languages. miniKanren
programs, especially relational interpreters, have been proven to be
useful in solving many problems by \citet{byrd2017unified}. A subtlety in 
writing miniKanren programs with large \scheme|conde| expressions, such as 
relational interpreters, is that the order of \scheme|conde| clauses sometimes 
affect the speed considerably. This phenomenon also appears when running 
miniKanren programs with the implementation in \citet{friedman_reasoned_2018}, 
one of the most well-understood implementation. This is because the search 
strategy of this implementation, interleaving depth-first search (\dfsi), gives 
left clauses higher ``search priority'' when a \scheme|conde| has more than two 
clauses. The situation is exacerbated as the number of clauses increases. This 
unfair treatment causes two problems when a \scheme|conde| expression has many 
clauses: the right-most clause can hardly contribute to query result; and 
programmers might need to tweak the clause order to maximize their programs' 
efficiency, depending on the distribution of queries.

The specification that every clause in the same \scheme|conde| is given equal 
``search priority'' is called fair \scheme|disj|. And search strategies with 
\emph{almost-fair} \scheme|disj| should give every clauss in the same 
\scheme|conde| similar priority. A related concept, fair \scheme|conj| is more 
complicated. We defer it to the next section.

To summarize our contribution, we
\begin{itemize}
    \item analyze an existing concept, fair \scheme|disj| by 
\citet{seres1999algebra}.
	\item propose a new concept, almost-fair \scheme|disj|.
	\item propose a new definition of fair \scheme|conj|.
	\item propose and implement balanced interleaving depth-first search 
	(DFS$_\textrm{bi}$), a new search strategy with almost-fair \scheme|disj|.
	\item propose and implement fair depth-first search (DFS$_\textrm{f}$), a 
	new search strategy with fair \scheme|disj|.
	\item implement in a new way breath-first search (BFS), a search strategy 
	with fair \scheme|disj| and fair \scheme|conj|
	(our code runs faster in all benchmarks and is simpler)
	\item prove our BFS implementation is equivalent with the one by 
	\citet{seres1999algebra}.
\end{itemize}

% A comparison of fairness of search strategies is in Fig.~\ref{fairness}.

% \begin{figure}[tbp]
% 	\begin{tabular}{|c|c|c|c|c|}
% 		\hline 
% 		fairness & DFS$_\textrm{i}$ & DFS$_\textrm{bi}$ & DFS$_\textrm{f}$ & BFS \\ 
% 		\hline 
% 		disj & unfair & almost-fair & fair & fair \\ 
% 		\hline 
% 		conj & unfair & unfair & unfair & fair \\ 
% 		\hline 
% 	\end{tabular} 
% 	\caption{fairness of search strategies}
% 	\label{fairness}
% \end{figure}

@@

Before going further into fairness, we would like to give a short review about 
state, search space, and goal, because fairness is defined in terms of them. 
Every answer given by miniKanren corresponds to a state. Each state records a 
way how some relations can be satisfied. In our case, the state is just a 
subsitution associating logic variables with terms. In more expressive 
miniKanrens (e.g. miniKanren with with symbolic constraints 
\citep{hemann2017framework}), states can include more information. A search 
space is a collection of states. And a goal is a function from a state to a 
search space. Goals check their inputs and produces possibly extended states. A 
goal might fail for some input, in which case the output search space would be 
empty.

Now we elaborate fairness by running queries about \scheme|repeato|, a 
relational definition that relates a term \scheme|x| with a non-empty list whose 
elements are \scheme|x| (Fig.~\ref{repeato}).

% 
% \begin{defn}[fair \scheme|conj|]
% A \scheme|conj| is fair iff it allocates computational resource evenly to 
% search spaces derived from states in the same bag, where bags are finite 
% list of states.
% \end{defn}

\begin{figure}
	\schemeinput{Figures/repeato.rkt}
	\caption{\texttt{repeat$^o$} and an example run}
	\label{repeato}
\end{figure}

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the answer. DFS$_\textrm{i}$, the current search
strategy, however, results in many more lists of \scheme|a|s than lists
of other letters. And some letters, e.g. \scheme|c| and \scheme|d|, are
rarely seen. The situation would be exacerbated if \scheme|conde| 
contains more clauses.

\begin{center}
    \begin{schemeregion}
    \schemeinput{Figures/repeato-disj-iDFS.rkt}
    \end{schemeregion}
\end{center}

Under the hood, the \scheme|conde| here is allocating computational effort to 
four trivially different search space. The unfair \scheme|disj| in 
\dfsi{} allocates much more effort to the first search space. On the contrary, 
fair \scheme|disj|s would allocate effort evenly to each search space.

\begin{center}
    \begin{schemeregion}
    \schemeinput{Figures/repeato-disj-DFSf.rkt}
    \end{schemeregion}
\end{center}

\begin{center}
    \begin{schemeregion}
    \schemeinput{Figures/repeato-disj-BFS.rkt}
    \end{schemeregion}
\end{center}

Running the same program again with almost-fair \scheme|disj| 
(e.g. DFS$_\textrm{bi}$) gives the same result. Almost-fair, however, is not 
completely fair, as shown by the following example.

\begin{center}
    \begin{schemeregion}
    \schemeinput{Figures/repeato-disj-biDFS.rkt}
    \end{schemeregion}
\end{center}

DFS$_\textrm{bi}$ is fair only when the number of goals is a power of 2, 
otherwise, some goals are allocated twice as many resources than the others. In 
the above example, where the \scheme|conde| has five clauses, the clauses of 
\scheme|b|, \scheme|c|, and \scheme|d| are allocated more resources.

We end this subsection with precise definitions of all levels of 
\scheme|disj| fairness.

Our definition of \emph{fair} \scheme|disj| is slightly generalize from the one 
by \citet{seres1999algebra}. Their definition is for binary disjunct. We 
generalize it to multi-arity one.

\begin{defn}[fair \scheme|disj|]
A \scheme|disj| is fair iff it allocates computational resource evenly to 
search spaces produced by goals in the same disjunct (e.g. clauses in the 
same 
\scheme|conde|).
\end{defn}

\begin{defn}[almost-fair \scheme|disj|]
A \scheme|disj| is almost-fair iff it allocates computational resource so 
evenly to search spaces produced by goals in the same disjunct that the maximal 
ratio of resources is bounded by a constant.
\end{defn}

\begin{defn}[unfair \scheme|disj|]
A \scheme|disj| is unfair iff it is not even almost-fair.
\end{defn}

@@

In the following program, the three \scheme|conde| clauses differ in a trivial way. So 
we expect lists of each letter constitute $1/4$ of the answer list. Search 
strategies with unfair \scheme|conj| (e.g. DFS$_\textrm{i}$, DFS$_\textrm{f}$), 
however, give us many more lists of \scheme|a|s than lists of other letters. 
And some letters (e.g. lists 
of \scheme|c|) are rarely found. Although DFS$_\textrm{i}$'s \scheme|disj| is unfair in general, 
it is fair when there is no call to relational definition in sub-goals, 
including this case. The situation would be worse if we add more \scheme|conde| 
clauses. The result with DFS$_\textrm{bi}$, whose \scheme|conj| is also unfair, is similar, but 
due to its different \scheme|disj|, the position of \scheme|b| and \scheme|c| are 
swapped.

\begin{center}
	\begin{schemeregion}
    \schemeinput{Figures/repeato-conj-iDFS.rkt}
	\end{schemeregion}
\end{center}

\begin{center}
	\begin{schemeregion}
    \schemeinput{Figures/repeato-conj-fDFS.rkt}
	\end{schemeregion}
\end{center}

Intuitively, search strategies with fair \scheme|conj| should produce each letter of 
lists equally frequently. Indeed, BFS does so.

\begin{center}
	\begin{schemeregion}
    \schemeinput{Figures/repeato-conj-fair.rkt}
	\end{schemeregion}
\end{center}

A more interesting situation is when the first conjunctive goal produces 
infinite many answers. Consider the following example, a naive specification of 
fair \scheme|conj| might require search strategies to produce all sorts of singleton 
lists, but no longer ones, which makes the strategies \emph{incomplete}.

\NOTE[MVC]{incomplete w.r.t. what? where's the definition of incomplete?}

\NOTE[LKC]{I am a bit confused. I assume completeness is a well-known concept 
in the context of logic programming. For example, this paper doesn't cite any 
source when it talks about completeness.

@@

\begin{center}
	\begin{schemeregion}
    \schemeinput{Figures/repeato-conj-infinite-naive.rkt}
	\end{schemeregion}
\end{center}

Our solution requires a search strategy with \emph{fair} \scheme|conj| to 
package answers in bags, where each bag contains finite answers, and to allocate 
resources evenly among search spaces derived from answers in the same bag. The 
way to package depends on search strategy. And how to allocate resources among 
search space related to different bags is unspecified. Our definition of fair 
\scheme|conj| is orthogonal with completeness. For example, a naively fair strategy 
is fair but not complete, while BFS is fair and complete. 
\NOTE[MVC]{Also here, complete w.r.t what?}

@@

\begin{center}
	\begin{schemeregion}
		\schemeinput{Figures/repeato-conj-infinite-sof.rkt}
	\end{schemeregion}
\end{center}

@@

\begin{figure}
	\schemeinput{Figures/balanced-disj.rkt}
	\caption{DFS$_\textrm{bi}$ implementation}
	\label{balanced-disj}
\end{figure}

Balanced interleaving DFS (DFS$_\textrm{bi}$) has almost-fair \scheme|disj| and unfair \scheme|conj|. 
The implementation of DFS$_\textrm{bi}$ differs from DFS$_\textrm{i}$ in the \scheme|disj| macro. We list the 
new \scheme|disj| with its helpers in Fig.~\ref{balanced-disj}. The first helper 
function, \scheme{split}, takes a list of goals \scheme{ls} and a procedure 
\scheme{k}, partitions \scheme{ls} into two sub-lists of roughly equal length, 
and returns the application of \scheme{k} to the two sub-lists. \scheme{disj*} 
takes a non-empty list of goals \scheme{gs} and returns a goal. With the help 
of \scheme{split}, it essentially constructs a \emph{balanced} binary tree 
where leaves are elements of \scheme{gs} and nodes are \scheme{disj2}s, hence 
the name of this search strategy. In contrast, the \scheme|disj| in DFS$_\textrm{i}$
constructs the binary tree with the same nodes but in the unbalanced form.

@@

\begin{figure}
	\schemeinput{Figures/fDFS.rkt}
	\caption{DFS$_\textrm{f}$ implementation}
	\label{fDFS}
\end{figure}

Fair DFS (DFS$_\textrm{f}$) has fair \scheme|disj| and unfair \scheme|conj|. The implementation of DFS$_\textrm{f}$ 
differs from DFS$_\textrm{i}$'s in \scheme{disj2} (Fig.~\ref{fDFS}). \scheme{disj2} is 
changed to call a new and fair version of \scheme{append-inf}. 
\scheme{append-inf/fair} immediately calls its helper, 
\scheme{append-inf/fair^}, with the first argument, \scheme{s?}, set to 
\scheme{#t}, which indicates that \scheme{s-inf} and \scheme{t-inf} haven't 
been swapped. The swapping happens at the third \scheme{cond} clause in the 
helper, where \scheme{s?} is updated accordingly. The first two \scheme{cond} 
clauses essentially copy the \scheme{car}s and stop recursion when one of the 
input spaces is obviously finite. The third clause, as we mentioned above, is 
just for swapping. When the fourth and last clause runs, we know that both 
\scheme{s-inf} and \scheme{t-inf} are ended with a thunk. In this case, a new 
thunk is constructed. The new thunk calls the driver recursively. Here changing 
the order of \scheme{t-inf} and \scheme{s-inf} won't hurt the fairness (though 
it will change the order of answers). We swapped them back so that answers are 
produced in a more natural order.

@@

BFS is fair in both \scheme|disj| and \scheme|conj|. Our implementation is based on 
DFS$_\textrm{f}$ (not DFS$_\textrm{i}$). All we have to do is apply two trivial changes to 
\scheme{append-map-inf}. First, rename it to \scheme{append-map-inf/fair}. 
Second, replace its use of \scheme{append-inf} to \scheme{append-inf/fair}.

The implementation can be improved in two ways. First, as mentioned in 
section 2.2, BFS puts answers in bags and answers of the same cost are in the 
same bag. In this implementation, however, it is unclear where this information 
is recorded. Second, 
\scheme{append-inf/fair} is extravagant in memory usage. It makes 
$O(n+m)$ new \scheme{cons} cells every time, where $n$ and $m$ are the 
``length''s of input search spaces. We address these issues in the first 
subsection.

@@

\begin{figure}
	\schemeinput{Figures/BFS-opt.rkt}	
	\caption{new and changed functions in optimized BFS that implements pure 
	features}
	\label{BFS-opt}
\end{figure}

@@

As mentioned in section 2.2, BFS puts answers in bags and answers of the 
same cost are in the same bag. The cost
information is recorded subtly -- the \scheme{car}s of a search space have cost 
0 (i.e. they are in the same bag), and the costs of answers in thunk are 
computed recursively then increased by one. It is even more subtle that
\scheme{append-inf/fair} and the \scheme{append-map-inf/fair} respects the cost 
information. We make these facts more obvious by changing the type of search 
space, modifying related function definitions, and introducing a few more 
functions.

The new type is a pair whose \scheme{car} is a list of answers (the bag), and 
whose \scheme{cdr} is either a \scheme{#f} or a thunk returning a search 
space. A falsy \scheme{cdr} means the search space is obviously finite.

Functions related to the pure subset are listed in Fig.~\ref{BFS-opt} (the 
others in Fig.~\ref{BFS-opt-cont}). They are compared with 
\citeauthor{seres1999algebra}'s implementation later. The first three functions 
in Fig.~\ref{BFS-opt} are search space constructors. \scheme{none} makes an 
empty search space; \scheme{unit} makes a space from one answer; and 
\scheme{step} makes a space from a thunk. The remaining functions do the same 
thing as before.

Luckily, the change in \scheme{append-inf/fair} also fixes the miserable space 
extravagance -- the use of \scheme{append} helps us to reuse the first bag of 
\scheme{t-inf}.

\citet{kiselyov2005backtracking} has shown that a \emph{MonadPlus} hides in 
implementations of logic programming system. Our BFS implementation is not an 
exception: \scheme{none}, \scheme{unit}, \scheme{append-map-inf}, and 
\scheme{append-inf} correspond to \scheme{mzero}, \scheme{unit}, \scheme{bind}, 
and \scheme{mplus} respectively.

\begin{figure}
	\schemeinput{Figures/BFS-opt-cont.rkt}	
	\caption{new and changed functions in optimized BFS that implements impure 
		features}
	\label{BFS-opt-cont}
\end{figure}

Functions implementing impure features are in Fig.~\ref{BFS-opt-cont}. The 
first function, \scheme{elim}, takes a space \scheme{s-inf} and two 
continuations \scheme{ks} and \scheme{kf}. When \scheme{s-inf} contains some 
answers, \scheme{ks} is called with the first answer and the rest space. 
Otherwise, \scheme{kf} is called with no argument. Here `s' and `f' means 
`succeed' and `fail' respectively. This function is an eliminator of search 
space, hence the name. The remaining functions do the same thing as before.

@@

We prove in Coq that the two BFSs are equivalent, i.e. \scheme{(run n g)} 
produces the same result (See supplements for the formal proof).
